
\newcommand{\etas}{\ensuremath{\eta_{\mathrm{s}}}}


\chapter{Introduction}

The three main forces involved in multi phase flows are the pressure forces, viscous forces and surface tension foces. Many techniques have been developed to simulate multiphase flows with surface tension. Smoothed Particle Hydrodynamics is a computational technique which was originally developed to model astrophysical problems \cite{Monaghan1977}, which was later extended to model a wide variety of problems in compuational physics. \cite{Monaghan1992, Monaghan1994}. SPH is a fully Lagrangian technique where fluid interfaces are advected with very little numerical diffusion. The SPH formalism readily accomodates extra physical effects and highly irregular, mobile or even deformable boundaries. However, SPH can often be more computationally expensive than competing methods for idealized problems. In this work we simulate surface tension acting on an interface between two fluids. 

\section{Methodology}

The SPH method was introduced in the late 1970's to solve the problems in astrophysics. There is no need of underlying grid for spatial discretization. As the SPh simulation is based on Lagrangian frame of reference, the fluid is discretized by moval spatial discretization points called particles. Only the initial domain is to be discretized and then these particles are moved according to their interactions. The physical propeerties like mass, density, velocity and internal energy are assigned to these particles. In accordance with the governing conservation equations, the fluid characteristics are calculated at each particle.

